---
title: "Predictive Analytics PS- 7"
author: "Shreyanshi Adak - 745"
date: "2026-03-19"
output: html_document
---

#Problem Set 7: Shrinkage Methods

Load Data and Preprocess
```{r}
library(ISLR)
library(glmnet)

# Load and clean data
data(Hitters)
Hitters <- na.omit(Hitters)

# Create model matrix (dummy variables)
x <- model.matrix(Salary ~ ., Hitters)[, -1]
y <- Hitters$Salary
```
Q1. Split the data set into training (remaining) and test set (100 observations)
```{r}
set.seed(123)
train_index <- sample(1:nrow(x), nrow(x) - 100)

x_train <- x[train_index, ]
y_train <- y[train_index]

x_test <- x[-train_index, ]
y_test <- y[-train_index]
```

Q2. Train least squares model and report train MSE and test MSE
```{r}
# Fit least squares
lm_fit <- lm(Salary ~ ., data = Hitters, subset = train_index)

# Predictions
train_pred <- predict(lm_fit, newdata = Hitters[train_index, ])
test_pred  <- predict(lm_fit, newdata = Hitters[-train_index, ])

# MSE
train_mse <- mean((y_train - train_pred)^2)
test_mse  <- mean((y_test - test_pred)^2)

train_mse
test_mse
```
Q3. Ridge Regression
```{r}
# Grid of lambda
lambda_grid <- 10^seq(10, -2, length = 100)

ridge_fit <- glmnet(x_train, y_train, alpha = 0, lambda = lambda_grid)

dim(coef(ridge_fit))
```
(ii) L2 norm for 50th and 60th lambda
```{r}
coef_50 <- coef(ridge_fit)[,50]
coef_60 <- coef(ridge_fit)[,60]

l2_50 <- sqrt(sum(coef_50[-1]^2))  # exclude intercept
l2_60 <- sqrt(sum(coef_60[-1]^2))

l2_50
l2_60
```
Comment: L2 norm decreases as λ increases → coefficients shrink.

(iv) Test MSE for λ₅₀ and λ₆₀
```{r}
pred_50 <- predict(ridge_fit, s = lambda_grid[50], newx = x_test)
pred_60 <- predict(ridge_fit, s = lambda_grid[60], newx = x_test)

mse_50 <- mean((y_test - pred_50)^2)
mse_60 <- mean((y_test - pred_60)^2)

mse_50
mse_60
```
Q4. Plot standardized coefficients vs λ
```{r}
plot(ridge_fit, xvar = "lambda", label = TRUE)
```

Interpretation:
As λ increases, coefficients shrink toward zero but never become exactly zero.

Q5. Plot vs ratio of L2 norms
```{r}
plot(ridge_fit, xvar = "norm", label = TRUE)
```

Interpretation:
Graph becomes smoother; x-axis now shows shrinkage relative to least squares instead of λ.

Q6. Ridge with Cross-Validation
```{r}
set.seed(123)
cv_ridge <- cv.glmnet(x_train, y_train, alpha = 0)

best_lambda <- cv_ridge$lambda.min

# Test MSE
ridge_pred <- predict(cv_ridge, s = best_lambda, newx = x_test)
ridge_mse <- mean((y_test - ridge_pred)^2)

best_lambda
ridge_mse
```
Q7. LASSO with Cross-Validation
```{r}
set.seed(123)
cv_lasso <- cv.glmnet(x_train, y_train, alpha = 1)

lasso_pred <- predict(cv_lasso, s = "lambda.min", newx = x_test)
lasso_mse <- mean((y_test - lasso_pred)^2)

# Number of non-zero coefficients
lasso_coef <- coef(cv_lasso, s = "lambda.min")
nonzero <- sum(lasso_coef != 0) - 1  # exclude intercept

lasso_mse
nonzero
```
Q8. Elastic Net with Cross-Validation
```{r}
set.seed(123)
cv_enet <- cv.glmnet(x_train, y_train, alpha = 0.5)

enet_pred <- predict(cv_enet, s = "lambda.min", newx = x_test)
enet_mse <- mean((y_test - enet_pred)^2)

# Non-zero predictors
enet_coef <- coef(cv_enet, s = "lambda.min")
enet_nonzero <- rownames(enet_coef)[enet_coef[,1] != 0]

enet_mse
enet_nonzero
```

Q9. Comparison of Models
```{r}
train_mse
test_mse
ridge_mse
lasso_mse
enet_mse
```
Comment:

Least Squares: High variance → higher test error

Ridge: Reduces variance → improves prediction

LASSO: Performs variable selection → sparse model

Elastic Net: Balance of Ridge & LASSO

Conclusion:
Shrinkage methods (especially LASSO/Elastic Net) generally give better prediction accuracy for Salary.