---
title: "Predictive Analytics PS- 6"
author: "Shreyanshi Adak - 745"
date: "2026-03-18"
output: html_document
---

Consider the “Hitters” data. For loading the data you need to install ISLR library. We wish to predict a baseball player’s salary on the basis of various predictors. So first we need to choose an optimal model. 


```{r}
rm(list=ls())
install.packages("ISLR")
library(ISLR)
head(Hitters)
dim(Hitters)
?Hitters
```


1. Clean the data by removing all rows that have missing value in any vari able. You will be working with this clean data. 

```{r}

df=na.omit(Hitters)
dim(df)
```

2. Split the data randomly into train-test keeping 80% as training sample and remaining as test sample. 

```{r}
set.seed(123)
train.id=sample(1:263,210)
train.data=df[train.id,] #Training data
head(train.data)
test.data=df[-train.id,] #Test data
head(test.data)
dim(train.data);dim(test.data)
```


3. Perform the best subset selection on the training sample. While choosing the best model fit up to a 19 variable model. Choose the best model according to adjusted R squared, BIC , AIC and Cp values. In case the best model differs according to the selection criteria, report the best fitted model according to each selection criterion. 

```{r}
install.packages("leaps")
library(leaps)
best.subset=regsubsets(Salary~.,data=train.data,nvmax=19)
summary(best.subset)
#Adjusted R squared*/*
adjr2=summary(best.subset)$adjr2
adjr2
cp=summary(best.subset)$cp
cp
bic=summary(best.subset)$bic
bic
coef(best.subset, which.min(bic)) #Best BIC Model
coef(best.subset, which.min(cp)) #Best cp Model
coef(best.subset, which.max(adjr2)) #Best Adjusted R squared Model
```

4. Find the training MSE and test MSE for each of the best models selected according to different criterion as mentioned in 3 and compare their values. 

```{r}
m.cp=lm(Salary~AtBat+Hits+Walks+CAtBat+CRuns+CRBI+CWalks+Division+PutOuts+Assists+NewLeague,data=train.data)
m.bic=lm(Salary~Walks+CAtBat+CHits+CHmRun+Division,data=train.data)
m.adjr=lm(Salary~AtBat+Hits+Walks+CAtBat+CRuns+CRBI+CWalks+Division+PutOuts+Assists+Errors+NewLeague,data=train.data)
mse.train.cp=1/210*(sum((train.data$Salary-predict(m.cp))^2)) #Training MSE for cp
mse.train.bic=1/210*(sum((train.data$Salary-predict(m.bic))^2)) #Training MSE for bic
mse.train.adjr=1/210*(sum((train.data$Salary-predict(m.adjr))^2)) #Training MSE for adjR2
mse.train.bic
mse.train.adjr
mse.train.cp
y.hat.test.cp=predict(m.cp, newdata=test.data)
y.hat.test.bic=predict(m.bic, newdata=test.data)
y.hat.test.adjr=predict(m.adjr, newdata=test.data)
mse.test.cp=mean((test.data$Salary-y.hat.test.cp)^2)
mse.test.cp
mse.test.bic=mean((test.data$Salary-y.hat.test.bic)^2)
mse.test.bic
mse.test.adjr=mean((test.data$Salary-y.hat.test.adjr)^2)
mse.test.adjr
out.best=data.frame(Measure=c("Cp","BIC","AdjR2"),Training_MSE=c(mse.train.cp,mse.train.bic,mse.train.adjr),
                    Test_MSE=c(mse.test.cp,mse.test.bic,mse.test.adjr))
out.best
```

5. Repeat 3-4 using backward stepwise selection method. 

```{r}
backward.subset=regsubsets(Salary~.,data=train.data,nvmax=19,method="backward")
summary(backward.subset)
coef(backward.subset, which.min(summary(backward.subset)$bic)) #Best BIC Model
coef(backward.subset, which.min(summary(backward.subset)$cp)) #Best cp Model
coef(backward.subset, which.max(summary(backward.subset)$adjr2)) #Best Adjusted R squared Model

m.cp2=lm(Salary~AtBat+Hits+Walks+CAtBat+CHits+CHmRun+CRuns+CWalks+Division+PutOuts+Assists+Errors+NewLeague,data=train.data)
m.bic2=lm(Salary~Walks+CAtBat+CHits+CHmRun+Division,data=train.data)
m.adjr2=lm(Salary~AtBat+Hits+Walks+CAtBat+CHits+CHmRun+CRuns+CWalks+Division+PutOuts+Assists+Errors+NewLeague,data=train.data)

mse.train.cp2=1/210*(sum((train.data$Salary-predict(m.cp2))^2)) #Training MSE for cp
mse.train.bic2=1/210*(sum((train.data$Salary-predict(m.bic2))^2)) #Training MSE for bic
mse.train.adjr2=1/210*(sum((train.data$Salary-predict(m.adjr2))^2)) #Training MSE for adjR2

y.hat.test.cp2=predict(m.cp2, newdata=test.data)
y.hat.test.bic2=predict(m.bic2, newdata=test.data)
y.hat.test.adjr2=predict(m.adjr2, newdata=test.data)
mse.test.cp2=mean((test.data$Salary-y.hat.test.cp2)^2)
mse.test.bic2=mean((test.data$Salary-y.hat.test.bic2)^2)
mse.test.adjr2=mean((test.data$Salary-y.hat.test.adjr2)^2)

out.backward=data.frame(Measure=c("Cp","BIC","AdjR2"),Training_MSE=c(mse.train.cp2,mse.train.bic2,mse.train.adjr2),
                    Test_MSE=c(mse.test.cp2,mse.test.bic2,mse.test.adjr2))
out.backward
```


6. Repeat 3-4 using forward stepwise selection method. 

```{r}
forward.subset=regsubsets(Salary~.,data=train.data,nvmax=19,method="forward")
summary(forward.subset)
coef(forward.subset, which.min(summary(forward.subset)$bic)) #Best BIC Model
coef(forward.subset, which.min(summary(forward.subset)$cp)) #Best cp Model
coef(forward.subset, which.max(summary(forward.subset)$adjr2)) #Best Adjusted R squared Model

m.cp3=lm(Salary~Walks+CAtBat+CHits+CRuns+CRBI+CWalks+Division+PutOuts+Assists+Errors+NewLeague,data=train.data)
m.bic3=lm(Salary~Walks+CAtBat+CHits+CRuns++CRBI+CWalks+Division,data=train.data)
m.adjr3=lm(Salary~AtBat+Hits+Walks+Years+CAtBat++CHits+CRuns+CRBI+CWalks+Division+PutOuts+Assists+Errors+NewLeague,data=train.data)

mse.train.cp3=1/210*(sum((train.data$Salary-predict(m.cp3))^2)) #Training MSE for cp
mse.train.bic3=1/210*(sum((train.data$Salary-predict(m.bic3))^2)) #Training MSE for bic
mse.train.adjr3=1/210*(sum((train.data$Salary-predict(m.adjr3))^2)) #Training MSE for adjR2

y.hat.test.cp3=predict(m.cp3, newdata=test.data)
y.hat.test.bic3=predict(m.bic3, newdata=test.data)
y.hat.test.adjr3=predict(m.adjr3, newdata=test.data)
mse.test.cp3=mean((test.data$Salary-y.hat.test.cp3)^2)
mse.test.bic3=mean((test.data$Salary-y.hat.test.bic3)^2)
mse.test.adjr3=mean((test.data$Salary-y.hat.test.adjr3)^2)

out.forward=data.frame(Measure=c("Cp","BIC","AdjR2"),Training_MSE=c(mse.train.cp3,mse.train.bic3,mse.train.adjr3),
                        Test_MSE=c(mse.test.cp3,mse.test.bic3,mse.test.adjr3))
out.forward

final=list(Best_Subset=out.best,Forward_Subset=out.forward,Backward_Subset=out.backward)
final

```








